---
title: "STAT 545A Homework 2"
author: Junbin ZHANG
date: Sept 17, 2018
output: github_document
---

# Bring rectangular data in
```{r}
## load gapminder
library(gapminder)
## load tidyverse
library(tidyverse)
```

# Smell test the data

## Is it a data.frame, a matrix, a vector, a list?
## What is its class?

```{r}
## show the class of the data
class(gapminder)
```

It is a data.frame.
It belongs to three classes: tbl_df, tbl, and data.frame

## How many variables/columns

```{r}
## show the number of columns
ncol(gapminder)
```

## How many rows/observations?

```{r}
## show the number of rows
nrow(gapminder)
```

## Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?

**For data.frame only**, we can also use `length(gapminder)` to replace `ncol(gapminder)`. We can also use `dim(gapminder)` to get the numbers of columns and rows at the same time. To be more complicated, `str(gapminder)` and `summary(gapminder)` are also possible to (indirectly) show the numbers of columns and rows

However, for `length()`, as an example, `length()` and `ncol()` will return different things if the input is a matrix. For the others, sometimes it may not be necessary to get so much redundant information.

## What data type is each variable?
```{r}
## show the structure of gapminder, which contains data type of each variable
str(gapminder)
```

There are six variables:

- "country" as Factor
- "continent" as Factor
- "year" as int (Integer)
- "lifeExp" as num (double)
- "pop" as int (Integer)
- "gdpPercap" as num (double)

# Explore individual variables

## Explore categorical variable

There are two categorical variabels: "country" and "continent". Let's explore "country" as a categorical variable.

### What are possible values (or range, whichever is appropriate) of each variable?

```{r}
## print unique values in "country"
unique(gapminder$country)
```

It prints out there are 142 unique values in this column, which matches the fact that there are 142 levels in "country"

### What values are typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$country)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
summary(gapminder$country)
```

It seems like every unique value in "country" has the same frequency, which is 12.

```{r}
## print the barplot of each unique value
barplot(table(gapminder$country))
```

It is not surprise that when we plot the frequency of every unique value in "country", each bar appears to be the same.

Therefore, we explore "continent" to check if the R code is correct.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$continent)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
summary(gapminder$continent)
```

So the most frequent (typical) value in "continent" is "Africa".

```{r}
## print the barplot of each unique value
barplot(table(gapminder$continent))
```

## Explore quantitative variable

There are four quantitative variabels: "year", "lifeExp", "pop" and "gdpPercap". Let's explore "year" as a quantitative variable.

### What are possible values (or range, whichever is appropriate) of each variable?

```{r}
## print the range of "year"
range(gapminder$year)
```

### What values are typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$year)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
table(gapminder$year)
```

It seems like every unique value in "year" has the same frequency, which is 142.

```{r}
## print the barplot of each unique value
barplot(table(gapminder$year))
```

Again, each bar in the barplot is the same.

Therefore, we explore "lifeExp" to check if the R code is correct.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$lifeExp)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
table(gapminder$lifeExp)
```

```{r}
## print the frequency of the most frequent value
tt[tt==max(tt)]
```

The most frequent value is 69.39, with 4 appearances.

```{r}
## print the barplot of each unique value
barplot(table(gapminder$lifeExp))
```

# Explore various plot types
# Use `filter()`, `select()` and `%>`

I would like to combine there two sections together to print some plots using piping and dplyr functions.

## A scatterplot of two quantitative variables.

Let's show the population (in log) in different years.

```{r}
sp <- gapminder %>% 
  # year as x axis and pop as y axis
  ggplot(aes(x=year, y=pop)) +
  # scale y axis by log10
  scale_y_log10() +
  # make it a scatterplot, and add transparancy
  geom_point(alpha=0.1)
# display scatterplot
sp
```

Let's also do a version with jitter.

```{r}
gapminder %>% 
  # year as x axis and pop as y axis
  ggplot(aes(x=year, y=pop)) +
  # scale y axis by log10
  scale_y_log10() +
  # make it a Jitter plot
  geom_jitter()
```


We then use facetting to show the trends in different continents. To make it clearer, we apply colors on different continents.

```{r}
sp +
  # show colors
  aes(color=continent) + 
  # facetting by continent
  facet_wrap(~ continent)
```

Let's select a single country (e.g. Canada), and show its trend of population throughout the years.

```{r}
gapminder %>% 
  # filter country as Canada
  filter(country == "Canada") %>% 
  # year as x axis and pop as y axis
  ggplot(aes(x=year, y=pop)) +
  # scale y axis by log10
  scale_y_log10() +
  # make it a scatterplot
  geom_point() +
  # try to draw a regression curve
  geom_smooth(method="lm", se=FALSE)
```


Let's see the difference of regression curve compared to line plot.


```{r}
gapminder %>% 
  # filter country as Canada
  filter(country == "Canada") %>% 
  # year as x axis and pop as y axis
  ggplot(aes(x=year, y=pop)) +
  # scale y axis by log10
  scale_y_log10() +
  # make it a lineplot with points
  geom_point() +
  geom_line()
```

The line plot only connects all data points instead of find out the relationship.

## A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.

In this section, we use "lifeExp" as the quantitative variable under analysis.

### A histogram of "lifeExp"

```{r}
gapminder %>% 
  # lifeExp as x axis
  ggplot(aes(x=lifeExp)) +
  # make it a histogram
  geom_histogram(bins=50, color="black")
```

### A densityplot of "lifeExp"

```{r}
gapminder %>% 
  # lifeExp as x axis
  ggplot(aes(x=lifeExp)) +
  # make it a densityplot
  geom_density(fill="grey")
```

Let's combine histogram and densityplot together
```{r}
gapminder %>% 
  # lifeExp as x axis
  ggplot(aes(x=lifeExp)) +
  # make it a histogram, also scale it similar to densityplot
  geom_histogram(bins=50, aes(y=..density..), color="black") +
  # make it a densityplot
  geom_density()
```

### A frequency polygon of "lifeExp"
```{r}
gapminder %>% 
  # lifeExp as x axis
  ggplot(aes(x=lifeExp)) +
  # make it a frequency polygon
  geom_freqpoly(bins=30)
```

## A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.

Let's show a boxplot for GDP in different continents. Though a `select()` does not affect the result, we use it just for a demonstration.

```{r}
gapminder %>% 
  # select gdpPercap and continent columns
  select(gdpPercap, continent) %>% 
  # continent as x axis and gdpPercap as y axis
  ggplot(aes(x=continent, y=gdpPercap)) +
  # scale y by log10, so the result is better
  scale_y_log10() +
  # make it a boxplot
  geom_boxplot()
```

Let's also make a violin plot.

```{r}
gapminder %>% 
  # select gdpPercap and continent columns
  select(gdpPercap, continent) %>% 
  # continent as x axis and gdpPercap as y axis
  ggplot(aes(x=continent, y=gdpPercap)) +
  # scale y by log10, so the result is better
  scale_y_log10() +
  # make it a violin plot
  geom_violin(fill="grey")
```

# But I want to do more!

## Evaluate this code and describe the result.

Let's run the code first to see what happens.
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

Here is a right version, and let's make a comparison.
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

The second one shows more results. That is because using `==`, country needs to be strictly equaled to a list with two string "Rwanda" and "Afghanistan", which does not match the data frame and leads to wrong results. While using `%in%`, country only needs to be either one of the two strings, and the results are complete.

## Use more of the dplyr functions for operating on a single table.

### `arrange()`

We can use `arrange()` to re-order rows.

```{r}
gapminder %>% 
  # re-order rows, first by continent, then country, then year
  arrange(continent, country, year) %>% 
  # re-order columns to first show continent then country
  select(continent, country, everything())
```

### `mutate()`

We can use `mutate()` to create new columns. For example, we can calulate log10 of "lifeExp", and show "pop" in million.

```{r}
gapminder %>% 
  # mutate lifeExp to lifeExpInLog10 and pop to popInMillion
  mutate(
    lifeExpInLog10 = log10(lifeExp),
    popInMillion = pop / 1000000
  ) %>% 
  # delete original columns
  select(-lifeExp, -pop)
```

### `summarise()` and `group_by`

We can use `summarise()` to calculate statistical summaries like mean values. We can also use `group_by()` to group same variables like continents.

```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(
    avg_gdpPercap = mean(gdpPercap),
    max_lifeExp = max(lifeExp),
    min_pop = min(pop),
    total = n()
  )
```

