---
title: "STAT 545A Homework 2"
author: Junbin ZHANG
date: Sept 17, 2018
output: github_document
---

# Bring rectangular data in
```{r}
## load gapminder
library(gapminder)
## load tidyverse
library(tidyverse)
```

# Smell test the data

## Is it a data.frame, a matrix, a vector, a list?
## What is its class?

```{r}
## show the class of the data
class(gapminder)
```

It is a data.frame.
It belongs to three classes: tbl_df, tbl, and data.frame

## How many variables/columns

```{r}
## show the number of columns
ncol(gapminder)
```

## How many rows/observations?

```{r}
## show the number of rows
nrow(gapminder)
```

## Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?

**For data.frame only**, we can also use `length(gapminder)` to replace `ncol(gapminder)`. We can also use `dim(gapminder)` to get the numbers of columns and rows at the same time. To be more complicated, `str(gapminder)` and `summary(gapminder)` are also possible to (indirectly) show the numbers of columns and rows

However, for `length()`, as an example, `length()` and `ncol()` will return different things if the input is a matrix. For the others, sometimes it may not be necessary to get so much redundant information.

## What data type is each variable?
```{r}
## show the structure of gapminder, which contains data type of each variable
str(gapminder)
```

There are six variables:

- "country" as Factor
- "continent" as Factor
- "year" as int (Integer)
- "lifeExp" as num (double)
- "pop" as int (Integer)
- "gdpPercap" as num (double)

# Explore individual variables

## Explore categorical variable

There are two categorical variabels: "country" and "continent". Let's explore "country" as a categorical variable.

### What are possible values (or range, whichever is appropriate) of each variable?

```{r}
## print unique values in "country"
unique(gapminder$country)
```

It prints out there are 142 unique values in this column, which matches the fact that there are 142 levels in "country"

### What values are typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$country)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
summary(gapminder$country)
```

It seems like every unique value in "country" has the same frequency, which is 12.

```{r}
## print the barplot of each unique value
barplot(table(gapminder$country))
```

It is not surprise that when we plot the frequency of every unique value in "country", each bar appears to be the same.

Therefore, we explore "continent" to check if the R code is correct.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$continent)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
summary(gapminder$continent)
```

So the most frequent (typical) value in "continent" is "Africa".

```{r}
## print the barplot of each unique value
barplot(table(gapminder$continent))
```

## Explore quantitative variable

There are four quantitative variabels: "year", "lifeExp", "pop" and "gdpPercap". Let's explore "year" as a quantitative variable.

### What are possible values (or range, whichever is appropriate) of each variable?

```{r}
## print the range of "year"
range(gapminder$year)
```

### What values are typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$year)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
table(gapminder$year)
```

It seems like every unique value in "year" has the same frequency, which is 142.

```{r}
## print the barplot of each unique value
barplot(table(gapminder$year))
```

Again, each bar in the barplot is the same.

Therefore, we explore "lifeExp" to check if the R code is correct.

```{r}
## print the most frequent value(s)
tt <- table(gapminder$lifeExp)
names(tt[tt==max(tt)])
```

```{r}
## print the frequency of each unique value
table(gapminder$lifeExp)
```

```{r}
## print the frequency of the most frequent value
tt[tt==max(tt)]
```

The most frequent value is 69.39, with 4 appearances.

```{r}
## print the barplot of each unique value
barplot(table(gapminder$lifeExp))
```

# Explore various plot types

[TBC]